---
title: "Reproducible Research: Peer Assessment 1"
author: "bdpnthr"
date: "`r Sys.Date()`"
output: html_document
---

## Loading and preprocessing the data
We first load the dataset and convert the date column to Date format.
```{r echo=TRUE}
activity <- read.csv("activity.csv", header=TRUE, stringsAsFactors=FALSE)
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
str(activity)
```

#What is mean total number of steps taken per day?
Calculate the total steps per day, plot a histogram, and compute the mean and median.
```{r echo=TRUE}
steps_per_day <- aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
hist(steps_per_day$steps, main="Total Steps per Day", xlab="Steps", col="lightblue")
mean_steps <- mean(steps_per_day$steps)
median_steps <- median(steps_per_day$steps)
mean_steps
median_steps
```

#What is the average daily activity pattern?
Examine the average number of steps per 5‑minute interval across all days.
```{r echo=TRUE}
avg_interval <- aggregate(steps ~ interval, data=activity, mean, na.rm=TRUE)
plot(avg_interval$interval, avg_interval$steps, type="l",
     main="Average Daily Activity Pattern", xlab="Interval", ylab="Average Steps")
max_interval <- avg_interval[which.max(avg_interval$steps), ]
max_interval
```

#Imputing missing values
Count missing values, replace them with the mean for that interval, and compare results.
```{r echo=TRUE}
missing_count <- sum(is.na(activity$steps))
missing_count

activity_imputed <- activity
for (i in 1:nrow(activity_imputed)) {
  if (is.na(activity_imputed$steps[i])) {
    interval_val <- activity_imputed$interval[i]
    activity_imputed$steps[i] <- avg_interval$steps[avg_interval$interval == interval_val]
  }
}

steps_per_day_imputed <- aggregate(steps ~ date, data=activity_imputed, sum)
hist(steps_per_day_imputed$steps, main="Total Steps per Day (Imputed)", xlab="Steps", col="lightgreen")
mean_steps_imputed <- mean(steps_per_day_imputed$steps)
median_steps_imputed <- median(steps_per_day_imputed$steps)
mean_steps_imputed
median_steps_imputed
```

#Are there differences in activity patterns between weekdays and weekends?
Create a factor variable for weekday/weekend and compare average activity patterns.
```{r echo=TRUE}
activity_imputed$daytype <- ifelse(weekdays(activity_imputed$date) %in% 
                                   c("Saturday","Sunday"), "weekend", "weekday")
activity_imputed$daytype <- factor(activity_imputed$daytype)

avg_daytype <- aggregate(steps ~ interval + daytype, data=activity_imputed, mean)

library(ggplot2)
ggplot(avg_daytype, aes(x=interval, y=steps)) +
  geom_line() +
  facet_wrap(~daytype, ncol=1) +
  labs(title="Weekday vs Weekend Activity Patterns",
       x="Interval", y="Average Steps")
```
